UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-03 ms/main.tex
added titles to captions

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{Colouring the projection by geographic coordinates.}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{Colouring the projection by phenotypic data}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{Using different colourings with the projection in figure~\ref{fig:umap_ukbb_pc10}, relationships between population structure, geography, and phenotypes become clear. (Top) Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. (Bottom) Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
    \label{fig:umap_ukbb_geo_map}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
   \caption{Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
    \label{fig:umap_world_map}
      \end{subfigure}
      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.}
\end{figure}

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.}
    \label{fig:supp_umap_ukbb_asia}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.}
    \label{fig:supp_umap_ukbb_car}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.}
    \label{fig:supp_umap_ukbb_eur}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_m}
\end{figure*}

%%%%% boxplots end
\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
    \end{subfigure}
    \caption{Alternate colouring of figre~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
    \label{fig:supp_umap_ukbb_worldmap_alt}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/admixture_plot_highlight_mountain_copy.pdf}
    \end{subfigure}
    \caption{Admixture plot of Hispanic individuals in the HRS. Those born in the Mountain census region fall between the white lines (indices 48 to 184)}
    \label{fig:supp_hrs_hisp_admix}
\end{figure*}

>>>>

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{Colouring the projection by geographic coordinates.}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{Colouring the projection by phenotypic data}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{Relationships between population structure, geography, and phenotype heterogeneity. (Top) Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. (Bottom) Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
    \label{fig:umap_ukbb_geo_map}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
   \caption{Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
    \label{fig:umap_world_map}
      \end{subfigure}
      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \sgcomment{Merge captions. Make sure to have overall title! }}
\end{figure}

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}. \sgcomment{Is this really a "material and method" comment? THe logical flow with the last comment is not clear to me.}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_asia}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_car}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_eur}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Eeosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_m}
\end{figure*}

%%%%% boxplots end
\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
    \end{subfigure}
    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
    \label{fig:supp_umap_ukbb_worldmap_alt}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/admixture_plot_highlight_mountain_copy.pdf}
    \end{subfigure}
    \caption{Admixture plot of Hispanic individuals in the HRS. Individuals born in the Mountain census region fall between the white lines (indices 48 to 184)}
    \label{fig:supp_hrs_hisp_admix}
\end{figure*}
